The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2024-01-29 paper.tex
property count corrections and other small improvements

Further axioms were introduced to precisely define certain categories.
In addition to an explicit textual definition of the category \enquote{Bone compound}, for example, we introduced the following axiom:
%
%\anno{BoneCompound}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{hasPart} some (\anno{Bone} or \anno{BoneCompound} or \anno{BonePart}))
\todo{KH: Ich bin nicht so der description logics Experte, Alex kannst du mal gucken ob ich das richtig formuliert habe?}
\begin{equation*}
\anno{BoneCompound} \sqsubseteq \exists \aurl{gfo}{hasPart}.(\anno{Bone} \sqcup \anno{BoneCompound} \sqcup \anno{BonePart})
\end{equation*}
%
This axiom states that a bone compound consists of further bone compounds, individual bones or bone parts.
Another example axiom postulates that phenotypes can be derived from anatomical entities or anatomical properties:
%
%\anno{Phenotype}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{derivedFrom} some (\anno{AnatomicalEntity} or \anno{AnatomicalProperty}))
\begin{equation*}
\anno{Phenotype} \sqsubseteq \exists \aurl{gfo}{derivedFrom}.(\anno{AnatomicalEntity} \sqcup \anno{AnatomicalProperty})
\end{equation*}
%
Finally, a spreadsheet template was developed for domain experts to specify domain-specific entities (ANNOds), strictly compliant with ANNOdc, see \cref{sec:annodc}.
This means that the domain experts had to insert more specific categories (single categories or category trees) under defined ANNOdc core categories and to describe them by suitable annotations and relations introduced in ANNOdc.

@@

The statistical information on the ANNO ontology is shown in Table 1.

\begin{table}[b]
\centering
\caption{Statistical information. Column values do not add up because some entities are defined in both subontologies.}
\label{tab:stats}
\begin{tabulary}{\textwidth}{lSSSSSSSSl}
\toprule
% Expressivity according to Protege: ANNOdc and combined: ALCHIQ(D), ANNOds: ALE
Ontology	&\textnormal{Classes}	&\textnormal{Object prop.}	&\textnormal{Data p.}	&\textnormal{Individuals}	&\textnormal{Annotation p.}	&\textnormal{Axioms}	&\textnormal{FMA links}				&\textnormal{TA links} &Expressivity\\
\midrule
ANNOdc		&41						&18								&1								&0							&19				&178					&11							&0					&OWL DL\\
ANNOds		&1166					&11								&0								&803						&43				&18231					&503						&206				&OWL DL\\
\midrule
combined	&1195					&21								&1								&803						&53				&18382					&514						&206				&OWL DL\\
\bottomrule
\end{tabulary}
\end{table}

>>>>

Further axioms were introduced to precisely define certain categories.
In addition to an explicit textual definition of the category \enquote{Bone compound}, for example, we introduced the following axiom:
%
%\anno{BoneCompound}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{hasPart} some (\anno{Bone} or \anno{BoneCompound} or \anno{BonePart}))
\begin{equation*}
\anno{BoneCompound} \sqsubseteq \exists \gfo{hasPart}.(\anno{Bone} \sqcup \anno{BoneCompound} \sqcup \anno{BonePart})
\end{equation*}
%
This axiom states that a bone compound consists of further bone compounds, individual bones or bone parts.
Another example axiom postulates that phenotypes can be derived from anatomical entities or anatomical properties:
%
%\anno{Phenotype}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{derivedFrom} some (\anno{AnatomicalEntity} or \anno{AnatomicalProperty}))
\begin{equation*}
\anno{Phenotype} \sqsubseteq \exists \gfo{derivedFrom}.(\anno{AnatomicalEntity} \sqcup \anno{AnatomicalProperty})
\end{equation*}
%
Finally, a spreadsheet template was developed for domain experts to specify domain-specific entities (ANNOds), strictly compliant with ANNOdc, see \cref{sec:annodc}.
This means that the domain experts had to insert more specific categories (single categories or category trees) under defined ANNOdc core categories and to describe them by suitable annotations and relations introduced in ANNOdc.

@@

The statistical information on the ANNO ontology is shown in \cref{tab:stats}.

\begin{table}[b]
\centering
\caption{Statistical information. Column values do not add up because some entities are defined in both subontologies.}
\label{tab:stats}
\begin{tabulary}{\textwidth}{lSSSSSSSSl}
\toprule
% Expressivity according to Protege: ANNOdc and combined: ALCHIQ(D), ANNOds: ALE
Ontology	&\textnormal{Classes}	&\textnormal{Object prop.}	&\textnormal{Data p.}	&\textnormal{Individuals}	&\textnormal{Annotation p.}	&\textnormal{Axioms}	&\textnormal{FMA links}				&\textnormal{TA links} &Expressivity\\
\midrule
ANNOdc		&41						&11								&1								&0							&19				&178					&11							&0					&OWL DL\\
ANNOds		&1166					&0								&0								&803						&43				&18231					&503						&206				&OWL DL\\
\midrule
combined	&1195					&11								&1								&803						&53				&18382					&514						&206				&OWL DL\\
\bottomrule
\end{tabulary}
\end{table}
